From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-29 paper.tex
Proof-reading by James -- Part 1

Building capable social agents requires to equip them with a range
of perceptual capabilities: while face and object tracking and
recognition, path planning, speech recognition or task learning are some of the current major research
directions, existing literature tends to evaluate each of these algorithms in
their own metric space, without considering the interaction quality at a global
level: Anzalone~\etal~\cite{anzalone} have recently argued in favor of
such a global evaluation, and they propose to assess these algorithms in term of
their capability to \emph{obtain the desired effect} in a human-robot interaction
context. They correspondingly propose metrics build around the measurement
of \emph{engagement} as indicator of the quality of the experience.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
with the current state of the interaction. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
\emph{connection} events such as directed gaze, mutual facial gaze is proposed. Not
relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
a determining factor for the assessment. Dinally, a recent study with
social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
non-verbal cues but they also underline possible limitations in long-term
scenarios.

The variety of these approaches reflects the fact that \emph{engagement} remains
a broad concept, fairly ill-defined and thus difficult to operationalize.
Therefore, instead of introducing ``yet another metric of engagement'', we
propose in this paper to introduce the more specific concept of
``\emph{with-me-ness}'': to what extend the human is ``with me'', the
robot, during the interaction.

We assess this level of ``with-me-ness'' by combining real-time estimation of the
attentional focus of the human with the expected, {\it a priori} focus
requested by the task, as estimated by the robot.

This article is organized in three sections: we first present a novel
method for on-line estimation of the focus of attention based on fast 6D head
pose estimation. We then propose to validate this approach with 
a real-world field experiment with children. We finally show how to compute
``with-me-ness'' over the course of the interaction to provide a new \emph{in-the-moment} measurement of
the quality of interaction, and compare levels computed by the robot with a
ground truth based on manual post-hoc annotations.

@@

The relation between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction, and the related gaze patterns, has been part of classic textbooks
like~\cite{argyle1969social} since decades. As such, there is little doubt that
measuring the direction of gaze is a useful proxy to estimate the (visual) focus
of attention of a social agent, and indeed this is one of the basic tool used in
social psychology.

Estimating attention using gaze is not new to robotics either. A
recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
way eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
performing such a measure on a robot, in real-time, and in ecologically
valid environments (which rules out bulky or invasive apparatus like
eye-trackers, or techniques requiring fine calibration and/or static
interactions) remains a challenge in HRI.

Looking at techniques that both operate on-line and have been deployed in field
experiments, one finds that most approaches rely on head pose estimation alone
(no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
and recent examples include~\cite{baxter2014tracking, anzalone}.

Approaches based on monocular 2D vision have been explored as
well~\cite{peters2010investigating}, with however limited robustness to
occlusions or lightning conditions, and over-reliance on tracking to maintain
real-time performances.  Our work relies on recent advances on template-based
face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) facial features extraction on 2D images, combined with 3D model
fitting, to obtain a fast, robust and stable 6D head pose estimate, that we
successfully deployed in field experiments involving child-robot interactions.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{field_of_attention}
    \caption{\small The visual field of attention is approximated to a
        40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
    pose intersect with this cone are considered \emph{in focus}.}
    \label{fig:vfoa}
\end{figure}

@@

We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
objects which lay inside this field of attention (Figure~\ref{fig:vfoa}). At a
given time, more than one object can therefore be \emph{in focus}.

Our implementation has two limitations: objects are approximated to points
(their origin), and we do not
check actual visibility: one object could be hidden by another, it would still
be considered as in focus. We did not address these limitations since our
experimental setup (involving relatively small objects with no occlusions) did
not mandate it. Techniques for more accurate assessment of the visual
perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.

@@

Our approach assumes that the pose of the objects of interest are available to
the system: as described in section~\ref{sec:system}, our implementation relies
on the ROS {\sc tf} framework to manage and make available to all software
modules the list of poses of existing objects (represented as {\it frames}), and
dedicated perception modules are in charged of publishing up-to-date
informations regarding the location of the objects of interest (the so-called
\emph{situation assessment}). Due to the nature of the experiment, most of the
points of interest considered for the experimental validation presented
hereafter are static with respect to the robot, thus simplifying the scene
perception.

@@

As presented above, we use the 6D head pose as an approximation of the
actual gaze direction, and we further approximate from here the participant's field of
attention. The assumption that such an double approximation of the field of attention
allows to derive the actual focus of attention needs to validated experimentally.
Our proposed experiment involves child-robot
interactions in the context of handwriting remediation.
This section details the experimental procedure and presents our results.

@@

@@

Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in
presence of a facilitator (one of the researchers).

@@

The subjects were typically located 50 cm away from the robot with the primary
(writing) tablet in front and the secondary one 30 cm to the left of the first
one.  The facilitator was located about 60 cm to the left of the subject.
Finally, two observers were located further away from the interaction field.
Figure~\ref{fig:setup} indicates accordingly the location of main areas of
interest (the two tablets, the robot, the facilitator and the observers).

The dependent variable is the measurement of the participants' VFoA, assessed in
terms of what are the attentional targets of the child over time. The face of the child
is acquired through a fixed webcam (Logitech {\sf c920}), placed on the table
(see Figure~\ref{fig:setup}), and the attentional targets are then computed as
presented in section~\ref{sec:vfoa}.

@@

Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
The study took place at school, in a separate room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number or a full
word), and presents a tactile tablet (equipped with a custom writing
application) to the robot; the robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pull back the tablet, corrects the robot's attempt by writing him/herself
on top or next to the robot's writing, and ``send'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until he/she
decides to train the robot on another word. In total, the children performed in
average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
and implementation of this experiment can be found in~\cite{Hood:2015}).

Besides, once per interaction, the robot interrupts the handwriting task to
propose to tell a story (taking about 2 min), and the turn-based
hand-writing task continues afterwards. The intended purpose of the
story-telling episode is to locally trigger a different set of attention behaviors.

@@

    \caption{\small ROS nodes involved in the VFoA estimation (yellow nodes were developed for this work).}
    \label{fig:system}
\end{figure}

The experiment was carried with an Aldebaran's {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent as individual frames
the possible point of interests: an object is considered to be in focus when its
frame lays within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

@@

Besides, we video-recorded the interactions, and performed a {\it post-hoc}
manual coding of the focus of attention by two annotators (Cronbach's
$\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
The manual coding forms our attentional \emph{ground-truth}.

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate and the
inter-judge agreement (Cronbach's $\alpha$).
%and the cosine similarity
%matrix.

>>>>

Building capable social agents requires to endow them with a range
of perceptual capabilities: however, while face and object tracking and
recognition, path planning, speech recognition or task learning are some of the current major research
directions, existing literature tends to evaluate each of these algorithms in
their own metric space, without considering the interaction quality at a global
level. Anzalone~\etal~\cite{anzalone} have recently argued in favor of
such a global evaluation, and they propose to assess these algorithms in term of
their capability to \emph{obtain the desired effect} in a human-robot interaction
context. They correspondingly propose metrics built around the measurement
of \emph{engagement} as indicator of the quality of the experience.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating non-verbal cues (gaze and smiles)
with the current state of the interaction in a Bayesian model. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
\emph{connection} events such as directed gaze, mutual facial gaze is proposed. Not
relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
a determining factor for the assessment. Finally, a recent study with
social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
non-verbal cues but they also underline possible limitations in long-term
scenarios.

The variety of these approaches reflects the fact that \emph{engagement} remains
a broad concept, fairly ill-defined and thus difficult to operationalize.
Therefore, instead of introducing ``yet another metric of engagement'', we
introduce the more specific concept of
\emph{with-me-ness}~\cite{sharma2014me}: to what extent the human is ``with me'', the
robot, during the interaction.

We measure \emph{with-me-ness} by comparing the
attentional focus of the human (as estimated in real-time by the robot) with the
expected, {\it a priori} targets of attention elicited by the task at hand.

This article is organized in three sections: we first present a novel
method for on-line estimation of the focus of attention based on fast 6D head
pose estimation. We then validate this approach with 
a real-world field experiment with children. We finally show how to compute
``with-me-ness'' over the course of the interaction to provide a new \emph{in-the-moment} measurement of
the quality of interaction, and compare levels computed by the robot with a
ground truth based on manual post-hoc annotations.

@@

The relation between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction, and the related gaze patterns, has been part of classic textbooks
like~\cite{argyle1969social} for decades. As such, there is little doubt that
measuring the direction of gaze is a useful proxy to estimate the (visual) focus
of attention of a social agent, and indeed this is one of the basic tools used in
social psychology.

Estimating attention using gaze is not new to robotics either. A
recent survey by Ruhland~\etal~\cite{ruhland2015review} gives in a broad
overview of eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}.
Performing such a measure on a robot, in real-time, and in ecologically
valid environments (which rules out bulky or invasive apparatus like
eye-trackers, or techniques requiring fine calibration and/or static
interactions) remains a challenge in HRI.

Looking at techniques that both operate on-line and have been deployed in field
experiments, one finds that most approaches rely on head pose estimation alone
(no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
Fanelli~\etal provides an overview of these approaches in~\cite{fanelli2012real},
and recent examples include~\cite{baxter2014tracking, anzalone}.

Approaches based on monocular 2D vision have been explored as
well~\cite{peters2010investigating}, with however limited robustness to
occlusions or lightning conditions, and over-reliance on tracking to maintain
real-time performances.  Our work relies on recent advances in template-based
face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) facial feature extraction on 2D images, combined with 3D model
fitting, to obtain a fast, robust and stable 6D head pose estimate, that we
successfully deployed in field experiments involving child-robot interactions.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{field_of_attention}
    \caption{\small Screenshot of the real-time attention estimation system. The visual field of attention is approximated to a
        40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
    pose intersect with this cone are considered \emph{in focus}.}
    \label{fig:vfoa}
\end{figure}

@@

We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At a
given time, more than one object can therefore be \emph{in focus}.

Our implementation has two limitations: objects are approximated to points
(they are considered in focus if their \emph{origin} lies in the field of
attention), and we do not
check actual visibility: one object could be hidden by another, it would still
be considered as in focus. We did not address these limitations since our
experimental setup (involving relatively small objects with no occlusions) did
not necessitate it. Techniques for more accurate assessment of the visual
perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.

@@

Our approach assumes that the pose of the objects of interest are available to
the system: as described in section~\ref{sec:system}, our implementation relies
on the ROS {\sc tf} framework to manage and make available to all software
modules the list of poses of existing objects (represented as {\it frames}), and
dedicated perception modules are in charge of publishing up-to-date
informations regarding the location of the objects of interest (the so-called
\emph{situation assessment}). Due to the nature of the experiment, most of the
points of interest considered for the experimental validation presented
hereafter are static with respect to the robot, thus simplifying the scene
perception.

@@

As presented above, we use the 6D head pose as an approximation of the
actual gaze direction, and we further approximate from here the participant's field of
attention. The assumption that such an approximation of the field of attention
allows to derive the actual focus of attention needs to validated experimentally.
Our proposed experiment involves child-robot
interactions in the context of handwriting remediation.
This section details the experimental procedure and presents our results.

@@

Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in the
presence of a facilitator (one of the researchers).

@@

@@

The subjects were typically located 50 cm away from the robot with the primary
(writing) tablet in front and the secondary one 30 cm to the left of the first
one.  The facilitator was located about 60 cm to the left of the subject.
Finally, two observers (visible by the child) were located further away from the interaction field.
Figure~\ref{fig:setup} indicates accordingly the location of main areas of
interest (the two tablets, the robot, the facilitator and the observers).

The dependent variable is the measurement of the participants' VFoA, assessed in
terms of what the attentional targets of the child are over time. The face of the child
is acquired through a fixed webcam (Logitech {\sf c920}), placed on the table
(see Figure~\ref{fig:setup}), and the attentional targets are then computed as
presented in section~\ref{sec:vfoa}.

@@

Six children (ages 5 to 6, 3 boys, 3 girls, none wearing glasses) were enrolled for this study.
The study took place at school, in an isolated room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number, or a full
word), and presents a tactile tablet (equipped with a custom writing
application) to the robot. The robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pulls back the tablet, corrects the robot's attempt by writing
on top of or next to the robot's writing, and ``sends'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until they
decide to train the robot on another word. In total, the children performed on
average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
and implementation of this experiment can be found in~\cite{Hood:2015}).

Once per interaction, the robot interrupts the handwriting task to
tell a story (taking about 2 min), and the turn-based
hand-writing task continues afterwards. The intended purpose of the
story-telling episode is to locally trigger a different set of attention
behaviors for the child.

@@

    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were developed for this work).}
    \label{fig:system}
\end{figure}

The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent the possible point
of interests as individual frames: an object is considered to be in focus when its
frame lies within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

@@

We video-recorded the interactions, and performed a {\it post-hoc}
manual coding of the focus of attention with two annotators (Cronbach's
$\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
The manual coding forms our attentional \emph{ground-truth}.

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate and the
inter-rater agreement (Cohen's $\kappa$).
%and the cosine similarity
%matrix.
